Comprehensive Algorithmic Resilience for \\Numeric Applications

### 2013-10-14 paper.tex
2013-10-13

%\documentclass[prodmode,acmtecs]{acmconf} % Aptara syntax
\documentclass{sig-alternate}
\usepackage{url}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{color}
\usepackage[ruled]{algorithm2e}

\newcommand{\sui}[1]{%
  \textcolor{green}{SC - #1}
}

@@

\title{Comprehensive Algorithmic Resilience for Scientific Applications }
\author{Sui Chen,
%\affil{Louisiana State University}
Greg Bronevetsky, Marc Casas,
%\affil{Lawrence Livermore National Laboratory}
Lu Peng
%\affil{Louisiana State University}
}

@@

Algorithmic resilience can be very complex to deploy in applications because a comprehensive resilience strategy must address all the types of errors that an application may experience.
This includes corruptions to numerical data as well as control information and data structures.
Full protection requires the combination of multiple techniques including algorithmic error detection, crash detection, replication of key control information as well as support for restart when errors are detected.
Further, it is necessary to configure the properties of the resilience mechanisms to make sure that they are cost-effectively meeting the user's accuracy and confidence targets.
This paper shows how all these issues can be resolved comprehensively in the context of three numeric applications.

@@

The limitations of automated resilience techniques has motivated the need for algorithmic techniques that can detect and tolerate errors by taking advantage of application-specific properties and invariants~\cite{amg_abft:2012, robustification:2010, abft:1984}.
Although these techniques can be very efficient, most research work on such techniques has primarily focused on how their algorithmic invariants can be leveraged to enable accurate error detection.
While this goal is important, it ignores the full complexity of the problem since it focuses on only a subset of the errors and does not consider combinations of different resilience techniques to efficiently and effectively address all the different types of faults.
In reality a comprehensive solution must incorporate a range of additional techniques such the replication of key variables, detection of memory violations and localized rollback on detected errors, as well as techniques to configure these various resilience methods to achieve the best performance given the user's accuracy and confidence targets.
%, they also require manual effort to incorporate into applications and may require deep insight into an application's invariants to deploy effectively.
%Fortunately, because real-world applications spend most of their execution is spent in a few application modules, most soft faults will occur in those modules rather than uniformly throughout the application.
%In scientific applications these modules are usually various types of libraries, including numeric solvers and domain-specific scientific packages.
%This indicates that if by focusing algorithmic resilience efforts on the key library routines, while leaving other modules unprotected or protected via more expensive mechanisms, it is possible to fully protect an application at a low cost.

This paper presents a comprehensive algorithmic resilience approach that considers all of the above issues.
We demonstrate it on different types of numerical applications: the Alternating Direction Method of Multipliers (ADDR) for Lasso problems~\cite{lasso:2011}, the Hattrick gravitational simulation~\cite{hattrick:2012} and the Digital Room Correction (DRC) acoustic correction application~\cite{drc:2012}.
We show how these three applications can be comprehensively protected from soft faults by adding three different resilience mechanisms to the routines where they spend the most time: algorithmic error checks, replication of critical data structures and checkpoint-restart of individual routines.
We demonstrate via fault injection experiments that although none of these techniques can individually protect applications from soft faults, their combination is highly effective.
Further, we demonstrate how these techniques can be configured to offer the right level of protection at any given fault rate or input type.
More specifically, these configurations can be used to produce a wide range of overheads and result accuracy levels and to tune the probabilistic confidence level that the given overhead or accuracy target will be met.
%\greg{Summarize real results}
%\sui{A tighter algorithmic checker bounds can be enabled at low error rates to completely eliminate erroneous results at a lower overhead, while a more lenient algorithmic checker would allow the user to produce more acceptable results in face of a higher error rate (such as 1e-09)}
For the three considered scientific codes, we correct more than 50\% of segmentation faults and we reduce the error in the outputs by one order of magnitude. The overhead is always lower than 20\%.

@@

@@

\subsection{Alternating Direction Method of Multipliers (ADDR)}
\label{sec:apps:lasso}
The ADDR method solves under-unconstrained linear problems $Ax=b$ for $x$ ($A$ has fewer rows than columns) while minimizing the function $f(x)$.
This is an iterative algorithm uses 64-bit precision and that spends most of its time in the following linear algebra operations, matrix-matrix multiplication, matrix-vector multiplication, rank-k update and Cholesky factorization, using the implementations of these routines from the GNU Scientific Library~\cite{gsl:2011}.
We focus on a specific application that uses the ADDR method to solve Lasso problems, which are defined as minimizing the function $\frac{1}{2} \left|| Ax - b \right||_2^2 + \lambda*\left|| x \right||_1$.
It is parallelized using MPI and our experiments focus on 4-processors runs of ADDR which computes linear systems of size $\{40, 80, 120, ..., 4000\} \times 500$ (40 to 4000 measurements with 500 variables each) and in the experiments below are denoted $S40$, $S80$, ..., $S4000$.
The values in $A$ and $b$ are generated by sampling a normal distribution with a mean of 0.
According to our experiments, the running time of this solver is not dependent on the kind of distribution the inputs are sampled from.

@@

@@

In this paper we consider a comprehensive resilience strategy that combines various types of error detection and recovery.
Error detection techniques are discussed in Section~\ref{sec:res_tech:err_det}, with Section~\ref{sec:res_tech:err_det:algo} focusing on algorithmic techniques to detect errors in the numeric portions of application state, while Section~\ref{sec:res_tech:err_det:mem} focuses on protection of data structures.
Section~\ref{sec:res_tech:cr} then considers recovery techniques based on both checkpoint-restart, which offers lower-overhead but higher-latency recovery, as well as replication of key variables, which reduces performance but enables fast recovery.
Finally, Section~\ref{sec:res_tech:eval} presents an experimental evaluation of the performance and accuracy properties of the resilient versions of the major routines used by our target applications, under a wide range of inputs and error injection rates.

@@

\vspace{-5pt}
\paragraph{Matrix-matrix multiplication (MMM)}
The operation $A \cdot B$ is checked using a matrix vector multiplication (MVM), via the identity: $(A \cdot B) \cdot x = A \cdot (B \cdot x)$, where $x$ is an error-checking vector.
Since we set the check vector $x$ to be all 1s this is equivalent to both sides computing the sums of the columns of matrix $A \cdot B$ using a single MVM on the left and two MVMs on the right.
Most errors in the computation of $A \cdot B$ will affect the sums of its columns and will thus be detected.
This check is efficient because the checker is asymptotically faster than the MMM algorithm, with MVM taking $O(n^2)$ time and MMM $O(n^3)$ time.

@@

%Experiment 2: Measure the error in the outputs of the algorithms with and without the algorithmic detectors.
%First, show a probability distribution histogram for all the codes at one error rate, input size and algorithmic detection threshold.
%Then show a summary of the entire space using the same heatmaps that we're using for final application results.
%Put the input size and error rate on the y-axis and the detection threshold on the x-axis.
%The color in each box should be the fraction of entries in the output that have an error of a given magnitude.
%Show a separate plot for a few different magnitudes.
\subsubsection{Memory Fault Detection}
\label{sec:res_tech:err_det:mem}

Detecting errors in application data structures requires checkers dedicated for this purpose.
Depending on the algorithm's properties different error detectors are appropriate.
In our work we considered the use of access protection hardware and replication of key data.

Since modern systems use 64-bit addressing the range of values that can be assumed by a given pointer is much larger than the amount of memory that can ever be assigned to an application.
This means that there is a very high probability that the corruption of a given pointer will cause it to point to an unallocated address.
Once this address is accessed the resulting violation will be detected by the memory protection hardware and communicated to the application via a Segmentation Fault or Bus Error signal.
This mechanism is cheap but suffers from a long delay between an error and its detection.

An alternative mechanism is to replicate key application variables and to compare the values of the replicas on each access to the variable.
While more expensive, this approach is effective if the number of key variables is small and the cost of rolling back computation is high.

The low cost of hardware memory protection mechanisms means that they are deployed in protecting all our routines.
In contrast, replication is only used to protect the RK routine because it is more sensitive to soft errors than the others and the low detection latency provided by replication ensures that any state corruptions are quickly corrected and have little impact on execution time.
This is important for resilience as well as performance since slowdowns due to frequent rollbacks increase the total number of errors the application is exposed to.

\subsection{Recovery via Checkpoint-Restart and \\Replication}
\label{sec:res_tech:cr}

@@

\begin{figure}[ht!]
%\vspace{-20pt}
\centering
\includegraphics[width=1.00\columnwidth]{figs/4_1_1_Exp1_linalg}
\includegraphics[width=1.00\columnwidth]{figs/4_1_1_Exp1_fft}
%\vspace{-5pt}
\caption{Overhead of resilience techniques when errors are not injected}
\label{fig:routine_detect_ovhd}
\end{figure}

%Experiment 1: Measure performance degradation of using error detectors as input sizes increase. In case of RK if we can run a version that doesn't do the multiple step sizes, do so but if this is difficult then don't worry about it.
We evaluate the costs of the above algorithmic error checkers by measuring their costs and effectiveness when running at different error rates on different types of inputs.
Figure \ref{fig:routine_detect_ovhd} measures the overheads of employing the algorithmic error detectors when no errors are injected, reporting the average of 100 runs.
The linear algebra routines were executed on $nxn$ square matrices where $n$ varies from 100 to 1000.
FFT ran on inputs vectors with 128K to 4M entries and RK ran on an integral with intervals $[0, t_1]$ where $t_1 \in \{10, 15, 20, 25, 30, 50, 75, 100\}$. The ODE system consists of a second-order nonlinear Van der Pol oscillator equation.
As our experiments show, the cost of algorithmic error detection is low and drops with increasing input sizes.

@@

Figure~\ref{fig:routine_all_ovhd} shows in the bars the overhead of all resilience techniques with no fault injection.
The data shows that checkpointing routine's input (input backup bar) consumes more time than algorithm specific data corruption mechanisms (checker bar). In all cases, the overheads decrease as the input set size increses.

@@

We now evaluate the effectiveness of our algorithmic detectors when errors are injected.
Figures~\ref{fig:algo_err_dist} show the fraction of entries in the outputs of each routine (y-axis) that have an error of a given magnitude (x-axis) as error magnitudes span from 1e-14 to 1e-4.
This figure is a focused view on few representative cases (each routine run on a single input size and error injection rate) across different resilience configurations: no error detection as well as three error detection thresholds.
As the error magnitude increases the fraction of output entries with errors of that magnitude drops.
Importantly, the fraction of erroneous entries is reduced at all magnitudes as a result of error detection and restart.
Further, this reduction is more significant as the detection threshold shrinks since smaller magnitude errors are detected.

Figure~\ref{fig:algo_err_heatmap} shows the same information but for all experimental configurations: (i) input sizes of n=100 to 500 for linear algebra, 512K-8M size inputs for FFT and 10, 20, 30, 50 and 100 for RK; (ii) error injection rates of 1e-7 to 1e-10.
The graph is a grid, with results for different routines listed vertically and different resilience configurations (no detection and different detection thresholds) listed horizontally.
The graph for each routine and resilience configuration presents the curves from Figure \ref{fig:algo_err_dist} as a heatmap.
The y-axis shows different routine input sizes and error injection rates, sorted according to the average number of error injections per run, in decreasing order from top to bottom.
The x-axis shows the different error magnitudes and the shade of each point is the fraction of the routine's entries that have a given error magnitude, with darker shades indicating more such entries.
Thus, each row of Figure~\ref{fig:algo_err_heatmap} represents the same information as each curve in Figure~\ref{fig:algo_err_dist}, with the shades of the squares representing the same information as the height of the line's points.

@@

The difference between RK routine and others motivates special discussion of its resilience properties.
Since RK's adaptive step size mechanism is already effective at tolerating many numerical errors we used it as the error detector for this routine without adjusting its tolerance parameters.
The configurable portions of the RK resilience method are thus the choice of checkpointing and/or replication strategy.
Configurations \texttt{Ckpt} and \texttt{NoChkpt} denote whether checkpointing was used and configurations \texttt{Re} and \texttt{NoRe} denote whether replication was used.
We consider checkpointing periods equal to 1, $10^2$, $10^3$, $10^4$ and $10^5$ calls to the RK4 stepping function and the following replication configurations: No replication, 1 pair of replicas for each variable in the source code and 1 pair of replicas for each occurrance of each variable in the source code (these configurations correspond to ``No repilcation'', ``Medium replication'' and ``Maximum replication'' in figure \ref{fig:rt_algo_err_heatmap}).

@@

Figure~\ref{fig:rk_routine_exp_exec} shows the RK's overhead when 0 up to 10 errors per run are injected.%, using the same format as Figure~\ref{fig:routine_exp_exec}.
Although the \texttt{NoRe\_NoCkpt} configuration has the lowest overhead when only a few errors are injected (low error rate or small execution times), as the number of errors per run increases this configuration quickly becomes extremely slow because an error is detected in almost every run.
Since each such detection requires a restart of the routine, at higher error rates this causes RK to restart many times before it completes.
In contrast, while the other configurations have a higher overhead for a few errors per run, their execution time scales much better with increasing error rates, making them significantly more practical.
By varying the checkpointing interval and the degree to which various RK variables are replicated it is possible to achieve a wide range of design points that provide the best performance at various error rates.

@@

Figure~\ref{fig:rt_algo_err_heatmap} shows the fraction of errors in RK's outputs for various checkpointing and replication configurations, in the same format as Figure~\ref{fig:algo_err_heatmap}.
It shows that the use of either resilience approaches reduces RK's vulnerability to soft errors, preserving the correctness of its output, and the use of both techniques is even more effective.
Figure~\ref{fig:rk_effectiveness} shows this from a different perspective: the fraction of RK runs where the root-mean-squared error of the output vector is below 1e-7 (y-axis), relative to the number of error injections per run (x-axis).
It shows that RK configurations with higher level of resilience are able to achieve this accuracy target more frequently.
This is primarily because by avoiding frequent rollbacks due to crashes these techniques reduce the total number of errors that are injected into the application run, which also reduces the probability that one such error will be missed by RK's algorithmic error detector.
As with overhead, the choice of checkpointing interval and replication degree make it possible to choose a wide range accuracy levels and probabilities that these targets will be satisfied.

@@

Having evaluated the resilience properties of the individual application routines and resilience techniques we now consider our three target applications ADDR, Hattrick and DRC.
Section~\ref{sec:eval:perf} reports on the impact of errors and resilience techniques on application performance. Section~\ref{sec:eval:acc} describes the distribution of errors in the results of these applications.
These results are based on a fault injection campaign that includes a range of fault injection rates, input sizes and configurations of resilience mechanisms, with 100 application runs for each combination of parameters.
This corresponds to a total of 66100 for lasso, 43200 for DRC, 102000 for Hattrick error injection experiments.
It was performed on compute nodes with two 2.33 GHz dual-core Xeon E5345 processors with 4GB Ram and RHEL 4 OS.

A given application run may terminate successfully or crash due to an unrecoverable error.
Unrecoverable errors occur when there is a segmentation fault in code outside our protected routines or if 5 rollbacks of a routine have been observed in a given run since such repeated rollbacks almost always indicate that the corruption in application state cannot be repaired.
Since we assume that the user needs application results regardless of the error rate, if the application crashes it is restarted from the beginning as many times as needed for it to complete.

To evaluate the behavior of these applications in a broad range of error environments we consider several error injection rates.
For reference, assuming a typical FIT of 100-100,000 for individual processing cores (FIT = failure in one billion hours of operation and this range has been experimentally observed for commodity electronics~\cite{mem_errors:2010}), an HPC system with one million cores will experience .1-100 errors per hour, which is equivalent to 3e-11 - 3e-14 errors per processor cycle on a 1Ghz processor.
The range of error injection rates used used in our experiments were chosen such that (i) on average one or more errors were injected in the executions of our target applications and (ii) at least some application runs completed without crashing.
Results for error rates that induce fewer than one error per run were extrapolated from our experiments with higher error rates.
This was done by considering only the runs where exactly one error was injected and trivially scaling the results to correspond to 1 in every $n$ runs being injected with an error while remaining $n-1$ runs complete with no injection.

%\greg{This is the section where we put all the pieces together and present the experimental results on full applications. We should describe our main use-case here: user wants to achieve a given accuracy with a given probability and we configure the algorithm to provide this at the lowest cost. QUESTION: how do we pick the parameters of each resilience technique? Which metric is being optimized?} \sui{Using the choices of those parameters derived from the experiments we have already performed (treat it as a ``training set'') and use those choices on some new experiment configurations (treat as ``test set''), and see what is the probability that those choices could satisfy the users' experiments.}

\subsection{Performance Impact}
\label{sec:eval:perf}

>>>>

%\documentclass[prodmode,acmtecs]{acmconf} % Aptara syntax
\documentclass{sig-alternate}
\usepackage{url}
\usepackage{ulem}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{color}
\usepackage[ruled]{algorithm2e}

\newcommand{\sui}[1]{%
  \textcolor{green}{SC-#1}
}

@@

\title{Comprehensive Algorithmic Resilience for \\Numeric Applications}
%\author{Sui Chen,
%\affil{Louisiana State University}
%Greg Bronevetsky, Marc Casas,
%\affil{Lawrence Livermore National Laboratory}
%Lu Peng
%\affil{Louisiana State University}
%}

@@

\sui{
Algorithmic resilience exploiting the invariants of certain algorithms is an old technique that has been around for many years. The invariants can serve as
correctness unit tests during development and turned off during run-time in favor of speed, but under the context of the upcoming exascale reliability challenges,
they can usually serve as useful fault checkers. However, unless other issues including input protection, control information
and critical data structure protection are resolved will algorithmic invariants help detecting and correcting errors.
% can be very complex to deploy in applications because a comprehensive resilience strategy must address all the types of errors that an application may experience.
%This includes corruptions to numerical data as well as control information and data structures.
%Full protection requires the combination of multiple techniques including algorithmic error detection, crash detection, replication of key control information as well as support for restart when errors are detected.
%Further, it is necessary to configure the properties of the resilience mechanisms to make sure that they are cost-effectively meeting the user's accuracy and confidence targets.
Further, certain design choices in exploiting algorithmic resilience imply trade-offs for cost-effectiveness of the final resilience strategy.
This paper presents our case study of exploiting algorithmic invariants for constructing comprehensive fault tolerance mechanisms and balancing performance and result accuracy in the context of large-scale computing.
}
%This paper shows how all these issues can be resolved comprehensively in the context of three numeric applications.

@@

The limitations of automated resilience techniques has motivated the need for algorithmic techniques that can detect and tolerate errors by taking advantage of application-specific properties and invariants~\cite{amg_abft:2012, robustification:2010, abft:1984}.
Although these techniques can be very efficient, most research work on such techniques has primarily focused on how their algorithmic invariants can be leveraged to enable accurate error detection.
While this goal is important, it ignores the full complexity of the problem since it focuses on only a subset of the errors \sui{(algorithmic correctness)} and 
does not consider combinations of different resilience techniques to efficiently and effectively address all the different types of faults.
In reality a comprehensive solution must incorporate a range of additional techniques such the replication of key variables, detection of memory violations and localized rollback on detected errors, as well as techniques to configure these various resilience methods to achieve the best performance given the user's accuracy and confidence targets.
%, they also require manual effort to incorporate into applications and may require deep insight into an application's invariants to deploy effectively.
%Fortunately, because real-world applications spend most of their execution is spent in a few application modules, most soft faults will occur in those modules rather than uniformly throughout the application.
%In scientific applications these modules are usually various types of libraries, including numeric solvers and domain-specific scientific packages.
%This indicates that if by focusing algorithmic resilience efforts on the key library routines, while leaving other modules unprotected or protected via more expensive mechanisms, it is possible to fully protect an application at a low cost.

This paper presents a comprehensive algorithmic resilience approach \sui{that combines existing checkpointing and algorithmic checker techniques that considers all of the above issues and demonstrate {\bf with fault injection experiments and statistical analyses} that such approach is able to significantly reduce the side effects of soft errors on parallel programs at a  slight performance cost.}
We demonstrate it on different types of numerical applications: the Alternating Direction Method of Multipliers (ADDR) for Lasso problems~\cite{lasso:2011}, the Hattrick gravitational simulation~\cite{hattrick:2012} and the Digital Room Correction (DRC) acoustic correction application~\cite{drc:2012}.
We show how these three applications can be comprehensively protected from soft faults by adding three different resilience mechanisms to the routines where they spend the most time: algorithmic error checks, replication of critical data structures and checkpoint-restart of individual routines.
We demonstrate via fault injection experiments that although none of these techniques can individually protect applications from soft faults, their combination is highly effective.
Further, we demonstrate how these techniques can be configured to offer the right level of protection at any given fault rate or input type.
More specifically, these configurations can be used to produce a wide range of overheads and result accuracy levels and to tune the probabilistic confidence level that the given overhead or accuracy target will be met.
%\greg{Summarize real results}
%\sui{A tighter algorithmic checker bounds can be enabled at low error rates to completely eliminate erroneous results at a lower overhead, while a more lenient algorithmic checker would allow the user to produce more acceptable results in face of a higher error rate (such as 1e-09)}
%For the three considered scientific codes, we correct more than 50\% of segmentation faults and we reduce the error in the outputs by one order of magnitude. The overhead is always lower than 20\%.

@@

\sui{As a metric for quantifying errors in output caused by soft faults, we perturb inputs to numeric applications ...}

@@

\subsection{Alternating Direction Method of Multipliers (ADDR)}
\label{sec:apps:lasso}
The ADDR method solves under-unconstrained linear problems $Ax=b$ for $x$ ($A$ has fewer rows than columns) while minimizing the function $f(x)$.
This is an iterative algorithm uses 64-bit precision and that spends most of its time in the following linear algebra operations, matrix-matrix multiplication, matrix-vector multiplication, rank-k update and Cholesky factorization, using the implementations of these routines from the GNU Scientific Library~\cite{gsl:2011}.
We focus on a specific application that uses the ADDR method to solve Lasso problems, which are defined as minimizing the function $\frac{1}{2} \left|| Ax - b \right||_2^2 + \lambda*\left|| x \right||_1$.
It is parallelized using MPI and our experiments focus on 4-processors runs of ADDR which computes linear systems of size $\{800, 1200, ..., 3600\} \times 500$ (800 to 3600 measurements with 500 variables each) and in the experiments below are denoted $S800$, $S1200$, ..., $S3600$.
The values in $A$ and $b$ are generated by sampling a normal distribution with a mean of 0.
According to our experiments, the running time of this solver is not dependent on the kind of distribution the inputs are sampled from.

@@

%\subsection{Barnes Hut}
%\label{sec:apps:barneshut}
%\sui{
%Barnes Hut is a parallel application for fast simulation of the motion of large amounts of bodies under the effects of gravity.
%It is designed to divide the cubic simulation volume using a quad-tree and approximating particles in distant cells
%by treating them as a large particle on the center of the cell's center of mass. 
%This dramatically reduces the number of interactions that needed to compute and reduces time complexity to $O(nlogn)$.
%When evaluating contribution of gravitational force on a certain point, a parameter $\theta$ dictates whether a cell is far enough from
%this point. Only when $\theta$ is larger than the quotient of the size of the cell and the distance from the cell's center of mass to the point ${\lvert \vec{s} \rvert} / {%\lvert \vec{d} \rvert}$ is the approximation used. Consequently, the algorithm is approximate and $\theta$ determines its accuracy. We use input ......
%}

@@

In this paper we consider a comprehensive resilience strategy that combines various types of error detection and recovery.
Error detection techniques are discussed in Section~\ref{sec:res_tech:err_det}, with Section~\ref{sec:res_tech:err_det:algo} focusing on algorithmic techniques to detect errors in the numeric portions of application state, while Section~\ref{sec:res_tech:err_det:mem} focuses on protection of data structures.
Section~\ref{sec:res_tech:cr} then considers recovery techniques based on both checkpoint-restart, which offers lower-overhead but higher-latency recovery, as well as replication of key pointers, which reduces performance but enables fast recovery.
Finally, Section~\ref{sec:res_tech:eval} presents an experimental evaluation of the performance and accuracy properties of the resilient versions of the major routines used by our target applications, under a wide range of inputs and error injection rates.

@@

\vspace{-5pt}
\paragraph{Matrix-matrix multiplication (MMM)}
The operation $A \cdot B$ is checked using a matrix vector multiplication (MVM), via the identity: $(A \cdot B) \cdot x = A \cdot (B \cdot x)$, where $x$ is an error-checking vector.
Since we set the check vector $x$ to be all 1s this is equivalent to both sides computing the sums of the columns of matrix $A \cdot B$ using a single MVM on the left and two MVMs on the right.
Most errors in the computation of $A \cdot B$ will affect the sums of its columns and will thus be detected.
This check is efficient because the checker is asymptotically faster than MMM, with MVM taking $O(n^2)$ time and MMM $O(n^3)$ time.

@@

%Experiment 2: Measure the error in the outputs of the algorithms with and without the algorithmic detectors.
%First, show a probability distribution histogram for all the codes at one error rate, input size and algorithmic detection threshold.
%Then show a summary of the entire space using the same heatmaps that we're using for final application results.
%Put the input size and error rate on the y-axis and the detection threshold on the x-axis.
%The color in each box should be the fraction of entries in the output that have an error of a given magnitude.
%Show a separate plot for a few different magnitudes.

% 4.1.2 is way too verbose, vague and not understandable
%\subsubsection{Memory Fault Detection}
%\label{sec:res_tech:err_det:mem}
%
%Detecting errors in application data structures requires checkers dedicated for this purpose.
%Depending on the algorithm's properties different error detectors are appropriate.
%In our work we considered the use of access protection hardware and replication of key data.
%
%Since modern systems use 64-bit addressing the range of values that can be assumed by a given pointer is much larger than the amount of memory that can ever be assigned to %an application.
%This means that there is a very high probability that the corruption of a given pointer will cause it to point to an unallocated address.
%Once this address is accessed the resulting violation will be detected by the memory protection hardware and communicated to the application via a Segmentation Fault or %Bus Error signal.
%This mechanism is cheap but suffers from a long delay between an error and its detection.
%
%An alternative mechanism is to replicate key application variables and to compare the values of the replicas on each access to the variable.
%While more expensive, this approach is effective if the number of key variables is small and the cost of rolling back computation is high.
%
%The low cost of hardware memory protection mechanisms means that they are deployed in protecting all our routines.
%In contrast, replication is only used to protect the RK routine because it is more sensitive to soft errors than the others and the low detection latency provided by %replication ensures that any state corruptions are quickly corrected and have little impact on execution time.
%This is important for resilience as well as performance since slowdowns due to frequent rollbacks increase the total number of errors the application is exposed to.

\subsection{Recovery via Checkpoint-Restart and \\Replication}
\label{sec:res_tech:cr}
Since bit-flip errors may corrupt the address of a load or store instruction and a bit-flip in a 64-bit address can easily become an invalid value pointing to
unallocated memory, a Segmentation Fault or Bus Error can be easily detected by the operating system.\\

@@

%\begin{figure}[ht!]
%\centering
%\includegraphics[width=1.00\columnwidth]{figs/4_1_1_Exp1_linalg}
%\includegraphics[width=1.00\columnwidth]{figs/4_1_1_Exp1_fft}
%\caption{Overhead of resilience techniques when errors are not injected}
%\label{fig:routine_detect_ovhd}
%\end{figure}

%Experiment 1: Measure performance degradation of using error detectors as input sizes increase. In case of RK if we can run a version that doesn't do the multiple step sizes, do so but if this is difficult then don't worry about it.
We evaluate the costs of the above algorithmic error checkers by measuring their costs and effectiveness when running at different error rates on different types of inputs.
%Figure \ref{fig:routine_detect_ovhd} measures the overheads of employing the algorithmic error detectors when no errors are injected, 
We report the averages of 100 runs.
The linear algebra routines were executed on $n \times n$ square matrices where $n$ varies from 100 to 1000.
FFT ran on inputs vectors with 128K to 16M entries and FIR on window widths of 1M to 8M.
RK ran on an integral with intervals $[0, t_1]$ where $t_1 \in \{10, 15, 20, 25, 30, 50, 75, 100\}$. 
Its ODE system is a second-order nonlinear Van der Pol oscillator equation.
Our experiments show that the cost of algorithmic error detection is low (< 10\%) for small and drops with increasing input sizes.

@@

Figure~\ref{fig:routine_all_ovhd} shows in the bars the overhead of all resilience techniques with no fault injection, separating into overheads due to algorithmic error detection, the checkpointing of routine inputs as well as checkpointing processor state.
The data shows that checkpointing the input of routines consumes more time than algorithmic error detection and that the cost of checkpointing processor state is negligible. 
In all cases, these overheads decrease as the input set size increases.

@@

We now evaluate the effectiveness of our algorithmic detectors when errors are injected.
In each case when an error is detected by our algorithmic detectors the routine's execution would restart from the beginning upto five restarts.
Figures~\ref{fig:algo_err_dist} show the fraction of entries in the outputs of each routine (y-axis) that have an error of a given magnitude (x-axis) as error magnitudes span from 1e-14 to 1e-4.
This figure is a focused view on few representative cases (each routine run on a single input size and error injection rate) across different resilience configurations: no error detection as well as three error detection thresholds.
As the error magnitude increases the fraction of output entries with errors of that magnitude drops.
Importantly, the fraction of erroneous entries is reduced at all magnitudes as a result of error detection and restart.
Further, this reduction is more significant as the detection threshold shrinks since smaller magnitude errors are detected.

Figure~\ref{fig:algo_err_heatmap} shows the same information but for all experimental configurations: (i) input sizes as above (ii) error injection rates of 1e-7 to 1e-10.
The graph is a grid, with results for different routines listed vertically and different resilience configurations (no detection and the different detection thresholds shown in the table at the bottom) listed horizontally.
The graph for each routine and resilience configuration presents the curves from Figure~\ref{fig:algo_err_dist} as a heatmap.
The y-axis shows different routine input sizes and error injection rates, sorted according to the average number of error injections per non-fault injected run, in decreasing order from top to bottom.
The x-axis shows the different error magnitudes and the shade of each point is the fraction of the routine's entries that have a given error magnitude, with darker shades indicating more such entries (each shade corresponds to a fraction that is some power of 10).
Thus, each row of Figure~\ref{fig:algo_err_heatmap} represents the same information as each curve in Figure~\ref{fig:algo_err_dist}, with the shades of the squares representing the same information as the height of the line's points.

The data shows that in general as the error detection threshold drops there are fewer entries that have an error of a given magnitude.
However, there are exceptions where the fraction of erroneous entries rises as the detection threshold shrinks (e.g. MMM and FT between NoFT and Threshold 1).
This is caused by the fact that we restart each routine on an error detection, which can causes the routine's execution to be exposed to additional error injections.
As the threshold is further tightened the amount of error in the output shrinks, indicating that error detection is ultimately effective if it is configured to be sufficiently accurate.
Another phenomenon that is observed is that as the number of errors injected in a given run decreases the degree of error in routine outputs also generally decreases.
However, this is not a fully consistent phenomenon, with some configurations that have more error injections being less vulnerable than others with a higher number of injections because the former had a higher injection rate but the latter a larger input size.
This indicates that while the number of injected errors is strongly correlated with overall application vulnerability to errors, other properties such as input size also play a role.

@@

The difference between RK routine and others motivates special discussion of its resilience properties.
Since RK's adaptive step size mechanism is already effective at tolerating many numerical errors we used it as the error detector for this routine without adjusting its tolerance parameters.
The configurable portions of the RK resilience method are thus the choice of checkpointing and/or replication strategy.
Configurations \texttt{Ckpt} and \texttt{NoChkpt} denote whether checkpointing was used and configurations \texttt{Re} and \texttt{NoRe} denote whether replication was used.
We consider checkpointing periods equal to 1, $10^2$, $10^3$, $10^4$ and $10^5$ calls to the RK stepping function.
Further, we consider the following replication configurations: (i) ``No replication'', (ii) `Medium replication'', where routine pointers are replicated but the replicas of each pointer are compared at a single code location where it is used, and (iii) ``Maximum replication'', where the comparison is performed on every pointer use.

@@

Figure~\ref{fig:rk_routine_exp_exec} shows the RK's overhead when between 0 and 10 errors per run are injected.%, using the same format as Figure~\ref{fig:routine_exp_exec}.
Although the \texttt{NoRe\_NoCkpt} configuration has the lowest overhead when only a few errors are injected (low error rate or small execution times), as the number of errors per run increases this configuration quickly becomes extremely slow because an error is detected in almost every run.
Since each such detection requires a restart of the routine, at higher error rates this causes RK to restart many times before it completes.
In contrast, while the other configurations have a higher overhead for a few errors per run, their execution time scales much better with increasing error rates, making them significantly more practical.
In particular, the overheads of the configurations with more replication and checkpointing scale better as the number of errors per run increase.
By varying the checkpointing interval and the degree to which various RK pointers are replicated it is possible to achieve a wide range of design points that provide the best performance at various error rates.

@@

Figure~\ref{fig:rt_algo_err_heatmap} shows the fraction of errors in RK's outputs for various checkpointing and replication configurations, in the same format and using the same experimental setup as in Figure~\ref{fig:algo_err_heatmap}.
It shows that while the use replication reduces the amount of error in RK's outputs, increased checkpointing doesn't have a significant effect on this accuracy metric.
Figure~\ref{fig:rk_effectiveness} evaluates the effect of checkpointing on accuracy from another perspective: the fraction of RK runs where the root-mean-squared error of the output vector is below 1e-7 (y-axis), relative to the number of error injections per run (x-axis).
It shows that RK configurations with higher level of resilience and in particular, more frequent checkpoints, are able to achieve this accuracy target more frequently.
This is primarily because by avoiding frequent rollbacks due to crashes these techniques reduce the total number of errors that are injected into the application run, which also reduces the probability that one such error will be missed by RK's algorithmic error detector.
As with overhead, the choice of checkpointing interval and replication degree make it possible to choose a wide range accuracy levels and probabilities that these targets will be satisfied.

@@

Having evaluated the resilience properties of the individual application routines and resilience techniques we now consider our three target applications ADDR, Hattrick and DRC.
Section~\ref{sec:eval:perf} reports on the impact of errors and resilience techniques on application performance. Section~\ref{sec:eval:acc} describes the distribution of errors in the results of these applications.
These results are based on a fault injection campaign that includes a range of fault injection rates, input sizes and configurations of resilience mechanisms, with 100 application runs for each combination of parameters.
This corresponds to a total of 66,100 fault injection experiments for ADDR (4-process parallel), 43,200 for DRC (sequential), 102,000 for Hattrick (sequential).
Our experiments were performed on compute nodes with two 2.33 GHz dual-core Xeon E5345 processors with 4GB Ram and RHEL 4 OS.

A given application run may terminate successfully or crash due to an unrecoverable error.
Unrecoverable errors occur when there is a segmentation fault in code outside our protected routines or if five rollbacks of a routine have been observed in a given run since such repeated rollbacks almost always indicate that the corruption in application state cannot be repaired.
Since we assume that the user needs application results regardless of the error rate, if the application crashes it is restarted from the beginning as many times as needed for it to complete.

To evaluate the behavior of these applications in a broad range of error environments we consider several error injection rates.
For reference, assuming a typical FIT of 100-100,000 for individual processing cores (FIT = failure in one billion hours of operation and this range has been experimentally observed for commodity electronics~\cite{mem_errors:2010, dram_error:2009}), an HPC system with one million cores will experience .1-100 errors per hour, which is equivalent to 3e-11 - 3e-14 errors per processor cycle on a 1Ghz processor. \sui{Hence, the experiments are referred to as ``accelerated experiments.''}
The range of error injection rates used used in our experiments were chosen such that (i) on average one or more errors were injected in the executions of our target applications and (ii) at least some application runs completed without crashing.
Results for error rates that induce fewer than one error per run were extrapolated from our experiments with higher error rates \sui{as described in \ref{sec:eval:extrapolation}}.
This was done by considering only the runs where exactly one error was injected and trivially scaling the results to correspond to 1 in every $n$ runs being injected with an error while remaining $n-1$ runs complete with no injection.\\

\subsection{Extrapolation Model}
\label{sec:eval:extrapolation}
\sui{
This section explains our modelling of massively parallel and long-running applications with the experiments mentioned in page \pageref{sec:eval}, which in nature
are accelerated experiments. We model larger scale experiments by extrapolating our experiments in the following ways:\\
}
